Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)

### 2015-08-04 docs/papers/shallow_performance/shallow_perf.tex
Add updated P8 results and some more discussion of them

\author{R.~W.~Ford, M.~Modani, G.~D.~Riley and A.~R.~Porter% <-this % stops a space
\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
\thanks{M.~Modani is with IBM.}
}

@@

\subsection{The `shallow' Program}
For this work we have used a benchmarking program that solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  In describing
our work, we term the version of the Shallow program obtained at the
beginning of this project the `original' version.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal version of shallow retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal version of shallow, we made the kernels
as fine-grained as possible. In this case, this resulted in eight
distinct kernels, each of which operated on a single field at a single
point (since we have chosen to use point-wise kernels). With a little
bit of tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
routine then consists of applying each of the kernels to all of the
points on the model mesh requiring an update. In its basic,
unoptimised (`vanilla') form, this PSy-layer routine then contains a
doubly-nested loop around each kernel call, as indicated by the
pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.

\begin{figure}
\centering
\includegraphics[width=85mm]{psykal_shallow}
\caption{A schematic of the vanilla \psykal version of the shallow code.}
\label{FIG_psykal_shallow_structure}
\end{figure}

@@

Our aim in this work is to achieve portable performance. Consequently,
we have performed tests with the hardware and compilers listed in
Table~\ref{TABLE_compilers}. Where a compiler is available on a given
piece of hardware, the version number used in this work is specified.
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around the
PowerPC xxx core with a clock speed of 4.1~GHz. The chip has xxx~MB of
last-level cache.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
                 & Gnu   & Intel       & Cray    & IBM     \\
\hline
Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
IBM Power 8      &       &             &         & 14.1     \\
\hline
\end{tabular}
\end{table}

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The compiler flags used in this work.}
\label{TABLE_compiler_flags}
\centering
\begin{tabular}{l|c}
\hline
Compiler  &  Flags \\
\hline
Gnu       & -Ofast -mtune=native -finline-limit=50000    \\
Intel     & -O3 -fast -fno-inline-factor    \\
Cray      & -O3 -O ipa5 -h wp               \\
IBM       & \bf{CHECK} -O5 -qinline=auto:level=10      \\
\hline
\end{tabular}
\end{table}

@@

In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
original version of the shallow code for the range of compilers and
hardware considered here. This summary demonstrates the effect of the
larger last-level cache of the Intel Ivy Bridge system compared to
that of the Haswell system; note the significant drop in performance
in going from a problem size of $256^{2}$ to a problem size of
$512^{2}$ for the black and green bars. The performance of the Ivy
Bridge system with the Cray or Intel compiler (red and purple bars)
only drops to this level when the domain size is increased to
$1024^{2}$. At this point, the working set no longer fits within cache
and performance is dominated by the bandwidth to main memory, making
it relatively insensitive to the choice of compiler.

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{orig_summary}
\caption{Summary of the performance of the original shallow code on 
the range of compilers and hardware under consideration.}
\label{FIG_orig_perf_summary}
\end{figure}

For this compiler-friendly form of the code, there is little
performance difference between the executables produced by the Gnu and
Intel compilers. However, the executable produced by the Cray compiler
generally performs significantly better. {\bf Why is this; can we work
out what it's doing that the others aren't?}

Note that the results for the Gnu compiler on Ivy Bridge show the same
drop-off (as problem size is increased) as those on the Haswell
system. This may be because Archer's login nodes seem to be running on
older hardware (E5-2650 with 20MB LLC) during the Archer upgrade. {\bf
  Need to re-run these cases once Archer is back up.}

Moving now the to the \psykal version of shallow,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest \psykal version for each of the compiler/CPU combinations. The
only significant difference between this and the plot of the
performance of the original version of the code in
Figure~\ref{FIG_orig_perf_summary} is that the difference between the
Gnu- and Intel-compiled executables is slightly more marked. This is
clarified in Figure~\ref{FIG_slowdown_summary} which plots the
percentage difference between the performance of the original and the
(best) \psykal versions of shallow for each compiler/CPU combination.
It is clear from this plot that the difference in the performance of
the Gnu- and Intel-compiled excutables on the Haswell system is due to
the Gnu compiler being unable to recover the performance achieved by
the original code. In contrast there is no such problem for the
Gnu-compiled executable on the Ivy Bridge system. We attribute this to
the more recent version of the compiler on the latter system (4.9.1 as
opposed to 4.8.2).

\begin{figure}[!t]
\centering
\includegraphics[width=7.5cm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal 
version of shallow with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

@@

In fact, for the Gnu and Cray compilers on the Ivy Bridge system, the
\psykal version of shallow is never more than 2\% slower than the
original and in some cases is faster.  This demonstrates that we can
reliably recover the performance of the original version of the code,
despite the significant restructuring required by the \psykal
approach.

Having show that, in general, we can recover and occasionally improve
upon the performance of the original version of shallow, the next
logical step is to examine the necessary code transformations.  We do
this for the $256^{2}$ case since this fits within cache on both the
Haswell and Ivy Bridge CPUs we are using here.
Table~\ref{TABLE_opt_breakdown} shows detailed timings for this case
after each transformation has been applied to the code. The same data
is visualised in Figure~\ref{FIG_opt_stages_256} where the performance
of the \psykal version for a given compiler/CPU is given relative to
the performance of the original with the same compiler/CPU.

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The mean time-per-step (s) for the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Compiler:          &  Gnu  & Intel   & Gnu & Intel & Cray & IBM \\
\hline
CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & Power8 \\
\hline
Original                               &  7.47E-04 & 7.06E-04 &	7.70E-04 & 7.77E-04 & 6.13E-04 &  1.17E-03  \\
Vanilla \psykal                        &  1.26E-02 & 8.64E-04 &	1.38E-02 & 9.82E-04 & 8.33E-04 &  7.87E-03  \\
Explicitly-sized arrays                &  1.26E-02 & 8.62E-04 &	1.38E-02 & 9.65E-04 & 6.96E-04 &  6.74E-03  \\
Kernels copied into module             &  1.10E-03 & 8.67E-04 &	9.42E-04 & 9.65E-04 & 7.10E-04 &  3.13E-03  \\
Loops fused                            &  1.09E-03 & 7.89E-04 &	9.07E-04 & 8.80E-04 & 7.35E-04 &  3.60E-03  \\
In-lined field-copy                    &  1.09E-03 & 7.90E-04 &	9.07E-04 & 8.73E-04 & 7.31E-04 &  3.62E-03  \\
Fused field-copy                       &  1.09E-03 & 9.25E-04 &	1.02E-03 & 1.04E-03 & 6.50E-04 &  4.11E-03  \\
In-line all kernels                    &  8.20E-04 & 7.14E-04 &	9.32E-04 & 8.52E-04 & 6.50E-04 &  1.07E-03  \\
Fully fuse 1st loop nest               &  8.10E-04 & 7.18E-04 &	7.76E-04 & 7.92E-04 & 6.23E-04 &  1.07E-03  \\
Explicit, constant array/loop bounds   &  8.05E-04 & 7.12E-04 &	7.81E-04 & 7.91E-04 & 6.13E-04 &  1.07E-03  \\
\hline
\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.8 & 1.8 & 0.03 & -9.0    \\
\hline
\end{tabular}
\end{table*}

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called has the most dramatic
effect on the performance of the compiled code; it now achieves almost
80\% of the performance of the original. The next most significant
transformation is that of (manually) in-lining the kernel code itself
into the bodies of the loops from which they are called. This gets the
\psykal version to within 8\% of the performance of the original.  The
dramatic performance improvement obtained from the first
transformation demonstrates that the Gnu compiler is unable to
optimise over separate files. Therefore, the fact that the
time-stepping loop in the \psykal version still retains a call to an
I/O routine and a call to the middle-layer, both of which are in
different files, is likely to account for this remaining performance
penalty of 8\%.

With the Intel compiler, the key transformations to recover
performance are different (second and fourth clusters in
Figure~\ref{FIG_opt_stages_256}).  Table~\ref{TABLE_opt_breakdown}
shows that even the vanilla \psykal version of shallow achieves some
80\% of the performance of the original version. Only two code
transformations were required to increase this to 100\%. The first of
these was the fusion of the computational loops. However, we found
that performance suffered (purple bar) when this was extended to
include the loop required for field copies at the end of a time-step
(Figure~\ref{FIG_time_smooth_code}). Further investigation of this
slow-down revealed that the compiler was unable to SIMD vectorise the
newly-fused loop after it had in-lined the body of the kernel called
from within it. Once this in-lining was done by hand (dark-blue bar),
the compiler was able to vectorise the loop and the performance of the
original version of shallow was recovered.

\begin{figure}[!t]
\centering
\includegraphics[width=95mm]{opt_stages_256}
\caption{The performance of the \psykal version of shallow for the
  $256^{2}$ domain at each stage of optimisation. Results are given as
  a percentage of the performance of the original code for each
  compiler/CPU combination. A figure greater than 100\% indicates that
  the \psykal version performs better than the original.}
\label{FIG_opt_stages_256}
\end{figure}

@@

! Copy new fields into current fields
DO J=1,N+1
  DO I=1,M+1
    Ufld(I,J) = UNEW(I,J)
  END DO
END DO
\end{verbatim}
\caption{Example of the coding structure that required manual
  in-lining of the kernel body (as indicated by commented-out lines)
  to retrieve the performance of the original version of shallow with
  the Intel compiler.}
\label{FIG_time_smooth_code}
\end{figure}

The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
evolution of the performance of the \psykal version of shallow with
the Cray compiler. Again, this has significant differences from the
behaviour seen with the Gnu and Intel compilers. As with the Intel
compiler, the performance of the Vanilla \psykal version is fairly
good at 74\% of that of the original. In contrast to the other
compilers consisdered so far, the first significant transformation is
simply to specify the bounds of the arrays being used within the
middle layer using module variables (as opposed to specifying them as
assumed size). This tells the compiler that all of the arrays used in
the middle layer have the same extent and also that all of the
computational loops are over almost all elements of these arrays. This
simple change gives a \psykal version that achieves 88\% of the
performance of the original.

@@

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.

%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals use top floats exclusively.
% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
% floats. This can be corrected via the \fnbelowfloat command of the
% stfloats package.

@@

>>>>

\author{R.~W.~Ford, M.~Modani, G.~D.~Riley and A.~R.~Porter% <-this % stops a space
\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
\thanks{M.~Modani is with IBM India.}
}

@@

\subsection{The `Shallow' Program}
For this work we have used a benchmarking program that solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourny~\cite{sadourny75}.
This software was originally written in 1984 by Paul Swarztrauber of
the National Center for Atmospheric Research, US.  However, in common
with many scientific codes, it has subsequently undergone some sort of
evolutionary development with subsequent people making various changes
and optimising it for previous generations of hardware.  In describing
our work, we term the version of the Shallow program obtained at the
beginning of this project the `original' version.

@@

Since any real oceanographic computational model must output results,
we ensure that any \psykal version of Shallow retains the Input/Output
capability of the original. This aids in limiting the optimisations
that can be performed on the \psykal version to those that should also
be applicable to full oceanographic models. Note that although we
retain the I/O functionality, all of the results presented in this work
carefully exclude the effects of I/O since it is compute performance
that interests us here.

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal version of Shallow, we made the kernels
as fine-grained as possible. In this case, this resulted in eight
distinct kernels, each of which operated on a single field at a single
point (since we have chosen to use point-wise kernels). With a little
bit of tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_shallow_structure}). The single PSy-layer
routine then consists of applying each of the kernels to all of the
points on the model mesh requiring an update. In its basic,
unoptimised (`vanilla') form, this PSy-layer routine then contains a
doubly-nested loop around each kernel call, as indicated by the
pseudo-code in Figure~\ref{FIG_psykal_shallow_structure}.

\begin{figure}
\centering
\includegraphics[width=85mm]{psykal_shallow}
\caption{A schematic of the vanilla \psykal version of the Shallow code.}
\label{FIG_psykal_shallow_structure}
\end{figure}

@@

Our aim in this work is to achieve portable performance. Consequently,
we have performed tests with the hardware and compilers listed in
Table~\ref{TABLE_compilers}. Where a compiler is available on a given
piece of hardware, the version number used in this work is specified.
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
'chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
is shared with all other chiplets so that there is a total of 96~MB of
last-level cache. The cores on the system we used had a clock speed of
2~GHz.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and CPUs used in this work. The use
  of a compiler on a given CPU is indicated by the specification of
  the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that CPU.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
                 & \multicolumn{4}{c|}{Compiler}             \\
\hline
                 & Gnu   & Intel       & Cray    & IBM     \\
\hline
Intel Haswell    & 4.8.2 & 14.0.0      &         &          \\
Intel Ivy Bridge & 4.9.1 & 14.0.1.106  & 8.3.3   &          \\
IBM Power 8      &       &             &         & 15.1.2     \\
\hline
\end{tabular}
\end{table}

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The compiler flags used in this work.}
\label{TABLE_compiler_flags}
\centering
\begin{tabular}{l|l}
\hline
Compiler  &  Flags \\
\hline
Gnu       & -Ofast -mtune=native -finline-limit=50000    \\
Intel     & -O3 -fast -fno-inline-factor    \\
Cray      & -O3 -O ipa5 -h wp               \\
IBM       & \parbox{5cm}{-O5 -qinline=auto:level=10\\ -qinline=auto:level=10 -qaltivec=be\\ -qdebug=NSIMDUBTHRESHOLD -qprefetch=dscr=0x1D7} \\
\hline
\end{tabular}
\end{table}

@@

In Figure~\ref{FIG_orig_perf_summary} we plot the performance of the
original version of the Shallow code for the range of compilers and
hardware considered here. This summary demonstrates the effect of the
larger last-level cache of the Intel Ivy Bridge system compared to
that of the Haswell system; note the significant drop in performance
in going from a problem size of $256^{2}$ to a problem size of
$512^{2}$ for the black and green bars. The performance of the Ivy
Bridge system with the Cray or Intel compiler (red and purple bars)
only drops to this level when the domain size is increased to
$1024^{2}$. At this point, the working set no longer fits within cache
and performance is dominated by the bandwidth to main memory, making
it relatively insensitive to the choice of compiler. The IBM Fortran
compiler for Linux on Power 8 is relatively immature and failed to
SIMD-vectorise two of the three major loop nests in the time-stepping
loop. Figure~\ref{FIG_orig_perf_summary} therefore also includes the
performance results obtained after the source had been modified to
work-around this limitation (we fissioned one loop nest and manually
unrolled another). Work is currently underway on updating the compiler
so that these modifications will be unecessary with a future release.

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{orig_summary}
\caption{Summary of the performance of the original Shallow code on
  the range of compilers and hardware under consideration. *After code
  modifications to encourage SIMD vectorisation (see text).}
\label{FIG_orig_perf_summary}
\end{figure}

For this compiler-friendly form of the code, there is little
performance difference between the executables produced by the Gnu and
Intel compilers. However, the executable produced by the Cray compiler
generally performs significantly better for all but the largest domain
size.

Moving now the to the \psykal version of Shallow,
Figure~\ref{FIG_best_psykal_perf_summary} plots the performance of the
fastest \psykal version for each of the compiler/CPU combinations. The
only significant difference between this and the plot of the
performance of the original version of the code in
Figure~\ref{FIG_orig_perf_summary} is that the difference between the
Gnu- and Intel-compiled executables is slightly more marked. This is
clarified in Figure~\ref{FIG_slowdown_summary} which plots the
percentage difference between the performance of the original and the
(best) \psykal versions of Shallow for each compiler/CPU combination.
It is clear from this plot that the difference in the performance of
the Gnu- and Intel-compiled excutables on the Haswell system is due to
the Gnu compiler being unable to recover the performance achieved by
the original code. In contrast there is no such problem for the
Gnu-compiled executable on the Ivy Bridge system. We attribute this to
the more recent version of the compiler on the latter system (4.9.1 as
opposed to 4.8.2).

\begin{figure}[!t]
\centering
\includegraphics[width=7.5cm]{best_psykal_summary}
\caption{Summary of the best performance achieved by any \psykal 
version of Shallow with each of the compilers and CPUs under 
consideration.}
\label{FIG_best_psykal_perf_summary}
\end{figure}

@@

In fact, for the Gnu and Cray compilers on the Ivy Bridge system, the
\psykal version of Shallow is never more than 2\% slower than the
original and in some cases is faster.  This demonstrates that we can
reliably recover the performance of the original version of the code,
despite the significant restructuring required by the \psykal
approach.

The most striking feature of Figure~\ref{FIG_slowdown_summary} is the
fact that the best \psykal version of Shallow is consistently faster
than the original on the Power 8. This confirms that this early
version of the IBM compiler struggled with the original form of the
Shallow source code.

Having shown that, in general, we can recover and occasionally improve
upon the performance of the original version of Shallow, the next
logical step is to examine the necessary code transformations.  We do
this for the $256^{2}$ case since this fits within last-level cache on
all of the CPUs we are using here.  Table~\ref{TABLE_opt_breakdown}
shows detailed timings for this case after each transformation has
been applied to the code. The same data is visualised in
Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal
version for a given compiler/CPU is given relative to the performance
of the original with the same compiler/CPU.

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The mean time-per-step (s) for the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Compiler:          &  Gnu  & Intel   & Gnu & Intel & Cray & IBM \\
\hline
CPU:               & \multicolumn{2}{c|}{Haswell} & \multicolumn{3}{c|}{Ivy Bridge} & Power8 \\
\hline
Original                               &  7.47E-04 & 7.06E-04 &	7.70E-04 & 7.77E-04 & 6.13E-04 &  1.17E-03  \\
Vanilla \psykal                        &  1.26E-02 & 8.64E-04 &	1.38E-02 & 9.82E-04 & 8.33E-04 &  7.87E-03  \\
Explicitly-sized arrays                &  1.26E-02 & 8.62E-04 &	1.38E-02 & 9.65E-04 & 6.96E-04 &  6.74E-03  \\
Kernels copied into module             &  1.10E-03 & 8.67E-04 &	9.42E-04 & 9.65E-04 & 7.10E-04 &  3.13E-03  \\
Loops fused                            &  1.09E-03 & 7.89E-04 &	9.07E-04 & 8.80E-04 & 7.35E-04 &  3.60E-03  \\
In-lined field-copy                    &  1.09E-03 & 7.90E-04 &	9.07E-04 & 8.73E-04 & 7.31E-04 &  3.62E-03  \\
Fused field-copy                       &  1.09E-03 & 9.25E-04 &	1.02E-03 & 1.04E-03 & 6.50E-04 &  4.11E-03  \\
In-line all kernels                    &  8.20E-04 & 7.14E-04 &	9.32E-04 & 8.52E-04 & 6.50E-04 &  1.07E-03  \\
Fully fuse 1st loop nest               &  8.10E-04 & 7.18E-04 &	7.76E-04 & 7.92E-04 & 6.23E-04 &  1.07E-03  \\
Explicit, constant array/loop bounds   &  8.05E-04 & 7.12E-04 &	7.81E-04 & 7.91E-04 & 6.13E-04 &  1.07E-03  \\
Fission new-field update loops         &           &          &          &          &          &            \\
Un-roll time-smoothing                 &           &          &          &          &          &            \\
\hline
\parbox{2.5cm}{\%-slow-down of best {\it c.f.} original} & 7.8 & 0.8 & 0.8 & 1.8 & 0.03 & -9.0    \\
\hline
\end{tabular}
\end{table*}

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
Shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called ('module in-lining') has
the most dramatic effect on the performance of the compiled code; it
now achieves almost 80\% of the performance of the original. The next
most significant transformation is that of manually in-lining the
kernel code itself into the bodies of the loops from which they are
called ('kernel in-lining'). This gets the \psykal version to within
8\% of the performance of the original.  The dramatic performance
improvement obtained from the first transformation demonstrates that
the Gnu compiler is unable to optimise over separate files. Therefore,
the fact that the time-stepping loop in the \psykal version still
retains a call to an I/O routine and a call to the middle-layer, both
of which are in different files, is likely to account for this
remaining performance penalty of 8\%.

With the Intel compiler, the key transformations to recover
performance are different (second and fourth clusters in
Figure~\ref{FIG_opt_stages_256}).  Table~\ref{TABLE_opt_breakdown}
shows that even the vanilla \psykal version of Shallow achieves some
80\% of the performance of the original version. Only two code
transformations were required to increase this to 100\%. The first of
these was the fusion of the computational loops. However, we found
that performance suffered (purple bar) when this was extended to
include the loop required for field copies at the end of a time-step
(Figure~\ref{FIG_time_smooth_code}). Further investigation of this
slow-down revealed that the compiler was unable to SIMD vectorise the
newly-fused loop after it had in-lined the body of the kernel called
from within it. Once this in-lining was done by hand (dark-blue bar),
the compiler was able to vectorise the loop and the performance of the
original version of Shallow was recovered.

\begin{figure}[!t]
\centering
\includegraphics[width=95mm]{opt_stages_256}
\caption{The performance of the \psykal version of Shallow for the
  $256^{2}$ domain at each stage of optimisation. Results are given as
  a percentage of the performance of the original code for each
  compiler/CPU combination. A figure greater than 100\% indicates that
  the \psykal version performs better than the original.}
\label{FIG_opt_stages_256}
\end{figure}

@@

! Copy new fields into current fields
DO J=1,N+1
  DO I=1,M+1
    Ufld(I,J) = UNEW(I,J)
  END DO
END DO
\end{verbatim}
\caption{Example of the coding structure that required manual
  in-lining of the kernel body (as indicated by commented-out lines)
  to retrieve the performance of the original version of Shallow with
  the Intel compiler.}
\label{FIG_time_smooth_code}
\end{figure}

The fifth cluster in Figure~\ref{FIG_opt_stages_256} shows the
evolution of the performance of the \psykal version of Shallow with
the Cray compiler. Again, this has significant differences from the
behaviour seen with the Gnu and Intel compilers. As with the Intel
compiler, the performance of the Vanilla \psykal version is fairly
good at 74\% of that of the original. In contrast to the other
compilers consisdered so far, the first significant transformation is
simply to specify the bounds of the arrays being used within the
middle layer using module variables (as opposed to specifying them as
assumed size). This tells the compiler that all of the arrays used in
the middle layer have the same extent and also that all of the
computational loops are over almost all elements of these arrays. This
simple change gives a \psykal version that achieves 88\% of the
performance of the original.

@@

As with the Gnu compiler, it is module in-lining and kernel in-lining
that are important for the IBM compiler (sixth cluster in
Figure~\ref{FIG_opt_stages_256}). Fusing loop nests actually harms
performance, just as it did for the Cray compiler.

The take-home message of all this is that while we can employ a
separation of concerns and recover performance, doing so is not
straightforward. Determining how best to optimise even a simple code
such as Shallow is highly compiler- and CPU-dependent. Decisions that
are good on one platform may be bad for another and if these are
written into the source code then they will accumulate over time and
will almost certainly result in a code that does not perform optimally
on any system.

@@

Some of our findings have been sent to the IBM compiler team and will
serve to improve future releases of the compiler.
